---
title: "Summary of sQTL and tuQTL associations on chr3 GWAS locus - B2 analysis"
author: "Silva Kasela"
date: "6/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path = here::here("covid_gwas", "freeze5", "chr3_coloc", "fig_summary_coloc_B2_other_molqtls/"), dev = c('png', 'pdf'))
options(knitr.table.format = "html") 

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(knitr)
library(kableExtra)

# ggplot2 theme
theme_set(theme_classic() +
            theme(plot.title = element_text(size = 14),
                  plot.subtitle = element_text(size = 12),
                  axis.title = element_text(size = 12),
                  axis.title.y = element_text(vjust = 2),
                  axis.title.x = element_text(vjust = -0.75),
                  axis.text = element_text(size = 11),
                  strip.text = element_text(size = 10),
                  legend.title = element_text(size = 12),
                  legend.text = element_text(size = 11)))

```

Can we detect colocalization signal between COVID-19 GWAS and sQTLs/tuQTLs?

```{r}
gwas_analysis <- "B2_ALL_leave_23andme"
lead_gwas <- "3:45823240:T:C"
qtl_analysis <- c("GTEx_v8_sqtl", "eQTL_catalogue_tx", "eQTL_catalogue_txrev")

```


Data

* COVID-19 HGI (Freeze 5 data): `r gwas_analysis` - hospitalized covid vs. population GWAS summary statistics
* eQTL Catalogue: transcript usage QTL (tuQTL) summary statistics from various (immune) cell types
* GTEx v8: sQTL summary statistics from 49 tissues

Approach

Run coloc between COVID-19 GWAS and sQTLs/tuQTLs, if there’s an sVariant/tuVariant within 100kb of the GWAS variant with nominal p-value < 10-4

* Coloc-standard
    + Run coloc in a 1Mb-region centered on the lead GWAS variant (+/- 500kb from the lead GWAS variant)
        * Priors used: `p1` = 1e-4, `p2` = 1e-4, `p12` = 5e-6
* Coloc-cond/mask
    + Run coloc in a 2Mb-region centered on the lead GWAS variant (+/- 1Mb from the lead GWAS variant)
    + Conditioning used for GTEx v8 with LD data from GTEx, masking used for eQTL catalogue data with LD from 1000G EUR population
        * `p1` = 1e-4, `p2` = 1e-4, `p12` = 5e-6
        * `mode` = 'iterative', `pthr` = 1e-04 (P-value threshold for secondary signals), `r2thr` = 0.01 (LD threshold for masking), `maxhits` = 3

```{r}
res_list <- list()
res_condmask_list <- list()

```

```{r read-in-data}
# eQTL Catalogue studies - eQTL Catalogue
tabix_paths <- read.table("https://raw.githubusercontent.com/eQTL-Catalogue/eQTL-Catalogue-resources/master/tabix/tabix_ftp_paths.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
tabix_paths <- tabix_paths %>%
  filter(!quant_method %in% c("ge", "microarray", "exon"))
tabix_paths$id <- sapply(tabix_paths$ftp_path, function(x) unlist(strsplit(x, "/"))[11])
tabix_paths$id <- sapply(tabix_paths$id, function(x) unlist(strsplit(x, "[.]"))[1])

# coloc-standard -------------------------
# Results of coloc
res_list <- lapply(qtl_analysis, function(x) {
  res <- read.table(here("covid_gwas", "freeze5", "chr3_coloc", "result", paste0("summary_coloc.", gwas_analysis, ".", x, ".txt")), header = T, sep = "\t", stringsAsFactors = F)
  if (grepl("eQTL_catalogue", x)) {
    res$id <- res$tuqtl_data
    res$id_match <- paste0(res$id, "_", res$tuqtl_gene_name, "_", res$tuqtl_molecular_trait_id)
    stopifnot(tabix_paths[paste0("eQTL_catalogue_", tabix_paths$quant_method) %in% x, "id"] %in% res$id)
    colnames(res) <- gsub("tuqtl", "eqtl", colnames(res))
  } else {
    res$id <- sapply(res$sqtl_data, function(x) sub("GTEx_v8_sqtl_splicing_", "", x))
    res$id_match <- paste0(res$id, res$sqtl_gene_name)
    colnames(res) <- gsub("sqtl", "eqtl", colnames(res))
  }
  return(res)
})
names(res_list) <- qtl_analysis

# coloc-condmask -------------------------
# Results of coloc
res_condmask_list <- lapply(qtl_analysis, function(x) {
  res <- read.table(here("covid_gwas", "freeze5", "chr3_coloc", "result_condmask", paste0("summary_coloc_condmask.", gwas_analysis, ".", x, ".txt")), header = T, sep = "\t", stringsAsFactors = F)
  if (grepl("eQTL_catalogue", x)) {
    res$id <- res$tuqtl_data
    res$id_match <- paste0(res$id, "_", res$tuqtl_gene_name, "_", res$tuqtl_molecular_trait_id)
    colnames(res) <- gsub("tuqtl", "eqtl", colnames(res))
  } else {
    res$id <- sapply(res$sqtl_data, function(x) sub("GTEx_v8_sqtl_splicing_", "", x))
    res$id_match <- paste0(res$id, res$sqtl_gene_name)
    colnames(res) <- gsub("sqtl", "eqtl", colnames(res))
  }

  # filter to take best probe = probe with min lead_gwas_variant_eqtl_pval => not needed
  # only one probe per gene in condmask files (i.e., only one probe per gene for coloc test anyway)
  id <- paste0(res$eqtl_data, "_", res$eqtl_gene_name, "_", res$eqtl_molecular_trait_id, "_", res$hit2)
  stopifnot(length(unique(id)) == length(id))
  rm(id)
  
  # filter to take best indep signal = hit2 with max pp4
  res <- res %>% 
    group_by(id_match) %>% 
    mutate(max_pp4 = max(PP.H4.abf)) %>%
    # arrange(desc(PP.H4.abf), .by_group = TRUE) %>% 
    ungroup() %>% 
    # distinct(eqtl_gene_name, eqtl_data, .keep_all = TRUE) %>% 
    select(-starts_with("coloc_")) %>% 
    data.frame()
  return(res)
})
names(res_condmask_list) <- qtl_analysis

```


## Summary of the results

#### P-value based lookup

What's the eQTL P-value of the lead sVariant/tuVariant and lead GWAS variant (`r lead_gwas`) in various studies from the eQTL Catalogue and GTEx?

```{r}
i <- 1
```

* `r names(res_list)[i]`

```{r fig-gwas-hit-eqtl-pval-heatmap-gtex-splicing, fig.width=12, fig.height=4}
res <- res_list[[i]]
plot_data <- res %>%
  select(eqtl_gene_name, id, lead_gwas_variant_eqtl_pval) %>%
  mutate(pval = -log10(lead_gwas_variant_eqtl_pval)) %>%
  pivot_wider(-lead_gwas_variant_eqtl_pval, names_from = id, values_from = pval) %>%
  tibble::column_to_rownames(var = "eqtl_gene_name") %>% 
  as.matrix()

pheatmap(mat = plot_data[c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"), ],
         scale = "none",
         color = viridis::viridis(100),
         na_col = "grey95",
         cluster_rows = F, cluster_cols = F,
         main = names(res_list)[i])

#ggplot(data = res, aes(x = eqtl_gene_name, y = eqtl_data, fill = -log10(lead_gwas_variant_eqtl_pval))) + 
#  geom_tile()

```

```{r}
i <- 2
```

* `r names(res_list)[i]`

```{r fig-gwas-hit-eqtl-pval-heatmap-eqtl-catalogue-tx, fig.width=14, fig.height=5}
res <- res_list[[i]]
plot_data <- res %>%
  select(eqtl_gene_name, id, lead_gwas_variant_eqtl_pval) %>%
  mutate(pval = -log10(lead_gwas_variant_eqtl_pval)) %>%
  pivot_wider(-lead_gwas_variant_eqtl_pval, names_from = id, values_from = pval) %>%
  tibble::column_to_rownames(var = "eqtl_gene_name") %>% 
  as.matrix()

pheatmap(mat = plot_data[c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"), ],
         scale = "none",
         color = viridis::viridis(100),
         na_col = "grey95",
         cluster_rows = F, cluster_cols = T,
         main = names(res_list)[i])

#ggplot(data = res, aes(x = eqtl_gene_name, y = eqtl_data, fill = -log10(lead_gwas_variant_eqtl_pval))) + 
#  geom_tile()

```


```{r}
i <- 3
```

* `r names(res_list)[i]`

```{r fig-gwas-hit-eqtl-pval-heatmap-eqtl-catalogue-txrev, fig.width=16, fig.height=5.5}
for (g in c("upstream", "contained", "downstream")) {
  res <- res_list[[i]]
  res <- res[grepl(g, res$eqtl_molecular_trait_id),]
  
  plot_data <- res %>%
    select(eqtl_gene_name, id, lead_gwas_variant_eqtl_pval) %>%
    mutate(pval = -log10(lead_gwas_variant_eqtl_pval)) %>%
    pivot_wider(-lead_gwas_variant_eqtl_pval, names_from = id, values_from = pval) %>%
    tibble::column_to_rownames(var = "eqtl_gene_name") %>% 
    as.matrix()
  
  # "contained" is missing CXCR6
  if (g == "contained") {
    plot_data <- rbind(plot_data, "CXCR6" = NA)
  }
  
  # ""downstream" is missing CXCR6
  if (g == "downstream") {
    plot_data <- rbind(plot_data, "XCR1" = NA)
  }
  
  pheatmap(mat = plot_data[c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"), ],
           scale = "none",
           color = viridis::viridis(100),
           na_col = "grey95",
           cluster_rows = F, cluster_cols = T,
           main = paste0(names(res_list)[i], " - ", g))
}

```

#### Summary of coloc-standard

```{r}
i <- 1
```

* `r names(res_list)[i]`

```{r fig-summary-coloc-gtex-splicing, fig.width=17, fig.height=12}
res <- res_list[[i]]

res %>%
  filter(!is.na(coloc_pp4)) %>% 
  pivot_longer(cols = starts_with("coloc_p"), names_to = "coloc_hyp", values_to = "coloc_pp") %>% 
  mutate(coloc_hyp = sub("coloc_pp", "H", coloc_hyp)) %>% 
  mutate(study = sub("_sqtl_splicing_", "\n", eqtl_data)) %>% 
  mutate(study = sub("_microarray_", "\n", study)) %>% 
  mutate(gene_name_f = factor(eqtl_gene_name, levels = c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"))) %>% 
  ggplot(aes(x = gene_name_f, y = coloc_pp, fill = coloc_hyp)) +
  labs(title = names(res_list)[i],
       x = "Gene name",
       y = "Posterior probability from coloc",
       fill = "Coloc hypothesis") +
  geom_col() +
  facet_wrap(~study, ncol = 7) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

```{r}
i <- 2
```

* `r names(res_list)[i]`

```{r fig-summary-coloc-eqtl-catalogue-tx, fig.width=17, fig.height=16}
res <- res_list[[i]]

res %>%
  filter(!is.na(coloc_pp4)) %>% 
  pivot_longer(cols = starts_with("coloc_p"), names_to = "coloc_hyp", values_to = "coloc_pp") %>% 
  mutate(coloc_hyp = sub("coloc_pp", "H", coloc_hyp)) %>% 
  mutate(study = sub("_tx_", "\n", eqtl_data)) %>% 
  mutate(gene_name_f = factor(eqtl_gene_name, levels = c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"))) %>% 
  ggplot(aes(x = gene_name_f, y = coloc_pp, fill = coloc_hyp)) +
  labs(title = names(res_list)[i],
       x = "Gene name",
       y = "Posterior probability from coloc",
       fill = "Coloc hypothesis") +
  geom_col() +
  facet_wrap(~study, ncol = 7) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

```{r}
i <- 3
```

* `r names(res_list)[i]`

```{r fig-summary-coloc-eqtl-catalogue-txrev, fig.width=17, fig.height=17}
for (g in c("upstream", "contained", "downstream")) {
  res <- res_list[[i]]
  res <- res[grepl(g, res$eqtl_molecular_trait_id),]
  
  p <- res %>%
    filter(!is.na(coloc_pp4)) %>% 
    pivot_longer(cols = starts_with("coloc_p"), names_to = "coloc_hyp", values_to = "coloc_pp") %>% 
    mutate(coloc_hyp = sub("coloc_pp", "H", coloc_hyp)) %>% 
    mutate(study = sub("_txrev_", "\n", eqtl_data)) %>% 
    mutate(gene_name_f = factor(eqtl_gene_name, levels = c("LIMD1", "SACM1L", "SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"))) %>% 
    ggplot(aes(x = gene_name_f, y = coloc_pp, fill = coloc_hyp)) +
    labs(title = paste0(names(res_list)[i], " - ", g),
         x = "Gene name",
         y = "Posterior probability from coloc",
         fill = "Coloc hypothesis") +
    geom_col() +
    facet_wrap(~study, ncol = 7) +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  print(p)
}
  
```

#### Summary coloc-condmask

__Compare coloc-standard and coloc-condmask results__

```{r fig-compare-coloc, fig.width=4.5, fig.height=4.5, out.width='33%'}
# GTEx - splicing -------------------
i = 1
stand <- res_list[[qtl_analysis[i]]]
stand <- stand[!is.na(stand$coloc_pp4),]
stopifnot(length(unique(stand$id_match)) == nrow(stand))

condmask <- res_condmask_list[[qtl_analysis[i]]] %>% 
  group_by(eqtl_data, eqtl_gene_name) %>% 
  arrange(desc(PP.H4.abf), .by_group = TRUE) %>% 
  ungroup() %>% 
  distinct(eqtl_gene_name, eqtl_data, .keep_all = TRUE) %>% 
  data.frame()
stopifnot(length(unique(condmask$id_match)) == nrow(condmask))

id <- intersect(stand$id_match, condmask$id_match)

plot(stand[match(id, stand$id_match), "coloc_pp4"], condmask[match(id, condmask$id_match), "PP.H4.abf"],
     xlab = "PP4 from coloc-standard", ylab = "Max PP4 from coloc-condmask",
     xlim = c(0, 1), ylim = c(0, 1),
     main = qtl_analysis[i])
abline(0, 1, col = "royalblue")

# eQTL Catalogue - tx -------------------
i = 2
stand <- res_list[[qtl_analysis[i]]]
stand <- stand[!is.na(stand$coloc_pp4),]
stopifnot(length(unique(stand$id_match)) == nrow(stand))

condmask <- res_condmask_list[[qtl_analysis[i]]] %>% 
  group_by(eqtl_data, eqtl_gene_name) %>% 
  arrange(desc(PP.H4.abf), .by_group = TRUE) %>% 
  ungroup() %>% 
  distinct(eqtl_gene_name, eqtl_data, .keep_all = TRUE) %>% 
  data.frame()
stopifnot(length(unique(condmask$id_match)) == nrow(condmask))

id <- intersect(stand$id_match, condmask$id_match)

plot(stand[match(id, stand$id_match), "coloc_pp4"], condmask[match(id, condmask$id_match), "PP.H4.abf"],
     xlab = "PP4 from coloc-standard", ylab = "Max PP4 from coloc-condmask",
     xlim = c(0, 1), ylim = c(0, 1),
     main = qtl_analysis[i])
abline(0, 1, col = "royalblue")

# eQTL Catalogue - txrev -------------------
i = 3
stand <- res_list[[qtl_analysis[i]]]
stand <- stand[!is.na(stand$coloc_pp4),]
stopifnot(length(unique(stand$id_match)) == nrow(stand))

condmask <- res_condmask_list[[qtl_analysis[i]]] %>% 
  group_by(id_match) %>% 
  arrange(desc(PP.H4.abf), .by_group = TRUE) %>% 
  ungroup() %>% 
  distinct(eqtl_gene_name, eqtl_data, id_match, .keep_all = TRUE) %>% 
  data.frame()
stopifnot(length(unique(condmask$id_match)) == nrow(condmask))

id <- intersect(stand$id_match, condmask$id_match)

plot(stand[match(id, stand$id_match), "coloc_pp4"], condmask[match(id, condmask$id_match), "PP.H4.abf"],
     xlab = "PP4 from coloc-standard", ylab = "Max PP4 from coloc-condmask",
     xlim = c(0, 1), ylim = c(0, 1),
     main = qtl_analysis[i])
abline(0, 1, col = "royalblue")

```

